---
title: "Basic model of evolutionary rescue in discrete time"
author: "Claudia Bank"
date: "2024-10-17"
output:
  html_document:
    df_print: paged
---

This notebook is meant to introduce you to simulations and calculations using R. In the following, we will simulate the evolutionary rescue model introduced in the lecture.

## Your tasks

Read through the code and execute it (using the "play" or "run" or "knit/preview" buttons - you might have to install some extra packages for the latter). Then "tackle" the following questions:

-   Do you understand what each line means? Which lines are (particularly) confusing?
-   Check out the questions following each code block and try to answer them.

## Troubleshooting

1.  If it all looks very confusing right now and you cannot answer any of the questions, do not despair! We will go through everything step by step during the practical and you will soon become familiar with such code. Note that we do **not** expect you to be able to answer all these questions today!
2.  If you cannot execute the code at all, tell us as soon as you can, so we can try to fix this together. Everyone should be able to run the code on their computer - it's not sufficient to peak onto your neighbor's screen.
3.  A great tool to understand what a line of code does is to print the output line by line, using the `print()` command. Be aware that in long loops this can create quite a mess of output.

------------------------------------------------------------------------

## Simulating evolutionary rescue in discrete time

In this notebook, we are implementing simulations of evolutionary rescue by a single mutation in discrete time, as discussed and analyzed in the work by Orr & Unckless. The model is as follows: a haploid population starts with a given number of wild-type and mutant individuals. Because of a sudden environmental change, the wild type is unfit and declines in the new environment. The mutant is beneficial and can survive and expand in the new environment, but it starts at a low (or zero) frequency in the total population. In order to rescue the population, it has to outcompete the wild type. If the mutant is not present at time 0, it has to appear through mutation and then spread in the population before extinction occurs.

## List of variables and parameters

-   $N_a$ current size of the wild-type (a) population (changes over time)
-   $N_A$ current size of the mutant (A) population (changes over time)
-   `N_init_a` initial population size of wild-type population (should be a positive integer)
-   `N_init_A` initial population size of the mutant population (should be a non-negative integer)
-   `decay_rate` detrimental effect of the environment on the wild type (i.e., the wild-type population decays at this rate; should be \>0 to induce potential extinction)
-   `sel_coeff` selective advantage of the mutant over the wild type (should be \>decay_rate to allow for evolutionary rescue)
-   `mut_rate` mutation rate from the wild type to the mutant genotype (should be \>0 and small)
-   `t_max` number of generations after which to stop the simulation (should be an integer \>0; this is not a model parameter but a setting for the simulations). In the code presented here, the simulation stops automatically if the population goes extinct, or if it recovers and exceeds 1.5 times the original total population size, so `t_max` can be set to a large value without creating extremely long simulation times.

## Simulate one generation of population size change

We want to observe the numbers of wild-type and mutant individuals over time. As first step, we simulate one generation of the population. At time $t$, the population has $N_a$ wild-type individuals and $N_A$ mutant individuals. First, these reproduce. This is modeled by drawing a Poisson random variable according to the expected absolute number $N_a \cdot (1-r)$ of wild-type individuals, where $r$ is the decay rate of the wild type, and $N_A \cdot (1-r+s)$ of mutant individuals, where $s$ is the selective advantage of the mutant over the wild time. Then mutation from $a$ to $A$ occurs as a Poisson random variable according to the given mutation rate. This defines the composition of the population at time $t+1$.

```{r}
simulate_one_gen <- function(N_a, N_A, decay_rate, sel_coeff, mut_rate) {
  # draw offspring according to Poisson distribution
  offsp_a <- rpois(1, N_a * (1-decay_rate))
  offsp_A <- rpois(1, N_A * (1-decay_rate+sel_coeff))
  # draw new mutants according to Poisson distribution
  mut_a_to_A <- rpois(1, offsp_a * mut_rate)
  
  # determine new population sizes of wild type and mutant
  N_a_new <- max(offsp_a - mut_a_to_A, 0)
  N_A_new <-  offsp_A + mut_a_to_A
  
  return(c(N_a_new, N_A_new))
}
# Test the function
print(simulate_one_gen(100,0,0.1,0.2,0.01))
```

### Questions

-   Why do we substract `mut_a_to_A` from `N_a_new` and not from `N_A_new`?
-   In the life cycle as defined above, do individuals reproduce first and then mutate, or vice versa?
-   Running the test output repeatedly, why do you almost always get mutant individuals in the next generation, even though you start without any mutants?
-   What does the function `c` do?

## Simulate a population trajectory

We can now track the dynamics of the population from one generation to another, but this is not enough. We want to follow the population from time 0 until it either is rescued or goes extinct. This is what the following function does.

## Create a periodic vector without any `if` functions
```{r}
# determine decay rates in good and bad times
decay_good_times <- 0
decay_bad_times <- 0.1
# determin duration of good and bad periods
period_good_times <- 7
period_bad_times <- 4
# determine the length of the simulation
t_max <- 100

# create a vector for one complete period
decay_vector_one_period <- c(rep(decay_good_times,period_good_times),rep(decay_bad_times,period_bad_times))
# create a vector with multiple periods, cut at length t_max
decay_vector_all <- head(rep(decay_vector_one_period,ceiling((t_max+1)/(period_bad_times+period_good_times))),t_max+1)

print(decay_vector_all)
```
Now turn this into a function:
```{r}
make_decay_vector <- function(decay_good_times,decay_bad_times,period_good_times,period_bad_times,t_max)
{
  # create a vector for one complete period
  decay_vector_one_period <- c(rep(decay_good_times,period_good_times),rep(decay_bad_times,period_bad_times))
  # create a vector with multiple periods, cut at length t_max
  return(head(rep(decay_vector_one_period,ceiling((t_max+1)/(period_bad_times+period_good_times))),t_max+1))
}

print(make_decay_vector(decay_good_times=0,decay_bad_times=0.1,period_good_times=3,period_bad_times=4,t_max=100))
```

## Simulate a trajectory given vectors of decay rates and selection coefficients
Now we write the output directly into a given data frame.
```{r}
gens <- 20
sel_vector <- rep(0.2,gens+1)
decay_vector <- make_decay_vector(decay_good_times=0,decay_bad_times=0.1,period_good_times=3,period_bad_times=4,t_max=gens)

# Note that the new function takes the tables as arguments, which naturally determine t_max (so it is not needed any more as an argument)
simulate_pop_table <- function(N_init_a, N_init_A, decay_vector, sel_vector, mut_rate) {
  # determine duration of the simulation
  times <- length(decay_vector)
  # send error message if vectors are ill-defined
  if (times != length(sel_vector))
    stop("The selection and decay vectors must have the same length.")
  # Create the table in which to save the results
  # generation counter
  gen <- 0:(times-1)
  # data frame in which to save the results
  pop_table <- data.frame(gen,decay_vector,sel_vector,N_a=NA,N_A=NA)
  # initiate the variables
  pop_table[1,4:5] <- c(N_init_a, N_init_A)
  # run the simulation
  for (i in 2:times) {
    # write the results directly and recursively into the table
    pop_table[i,4:5] <- simulate_one_gen(pop_table[i-1,4],pop_table[i-1,5],decay_vector[i], sel_vector[i], mut_rate)
    # condition to stop the simulation before times: either the population exceeds 10 times the original population size, or it goes extinct
    if (sum(pop_table[i,4:5])>=10*(N_init_a+ N_init_A) | sum(pop_table[i,4:5])==0) break
  }
  return(pop_table)	
}
 print(simulate_pop_table(N_init_a=10, N_init_A=0, decay_vector, sel_vector, mut_rate=0.1))
```

