---
title: "Basic model of evolutionary rescue in discrete time with immigration and competition"
author: "Claudia Bank"
date: "2024-10-22"
output:
  html_document:
    df_print: paged
---
This noteboook is derived from the Research Practical code "Basic model of evolutionary rescue in discrete time". Here, we introduce stochastic immigration and competition according to a logistic growth model.

------------------------------------------------------------------------

## Simulating evolutionary rescue in discrete time

In this notebook, we are implementing simulations of evolutionary rescue by a single mutation in discrete time, as discussed and analyzed in the work by Orr & Unckless. The model is as follows: a haploid population starts with a given number of wild-type and mutant individuals. Because of a sudden environmental change, the wild type is unfit and declines in the new environment. The mutant is beneficial and can survive and expand in the new environment, but it starts at a low (or zero) frequency in the total population. In order to rescue the population, it has to outcompete the wild type. If the mutant is not present at time 0, it has to appear through mutation and then spread in the population before extinction occurs.

## List of variables and parameters

-   $N_a$ current size of the wild-type (a) population (changes over time)
-   $N_A$ current size of the mutant (A) population (changes over time)
-   `N_init_a` initial population size of wild-type population (should be a positive integer)
-   `N_init_A` initial population size of the mutant population (should be a non-negative integer)
-   `decay_rate` detrimental effect of the environment on the wild type (i.e., the wild-type population decays at this rate; should be \>0 to induce potential extinction)
-   `sel_coeff` selective advantage of the mutant over the wild type (should be \>decay_rate to allow for evolutionary rescue)
-   `mut_rate` mutation rate from the wild type to the mutant genotype (should be \>0 and small)
- `mig_rate_a` is the immigration rate of "a" individuals
-   `t_max` number of generations after which to stop the simulation (should be an integer \>0; this is not a model parameter but a setting for the simulations). In the code presented here, the simulation stops automatically if the population goes extinct, or if it recovers and exceeds 1.5 times the original total population size, so `t_max` can be set to a large value without creating extremely long simulation times.

## Simulate one generation of population size change

We want to observe the numbers of wild-type and mutant individuals over time. As first step, we simulate one generation of the population. At time $t$, the population has $N_a$ wild-type individuals and $N_A$ mutant individuals. First, these reproduce. This is modeled by drawing a Poisson random variable according to the expected absolute number $N_a \cdot (1-r)$ of wild-type individuals, where $r$ is the decay rate of the wild type, and $N_A \cdot (1-r+s)$ of mutant individuals, where $s$ is the selective advantage of the mutant over the wild time. Then mutation from $a$ to $A$ occurs as a Poisson random variable according to the given mutation rate. This defines the composition of the population at time $t+1$.

## Include competition with logistic growth and migration
Steps of the life cycle:

*adult migrants of type $a$ arrive
*reproduction: $a$ individuals always suffer from the decay rate, independent of the current population size; the growth rate of $A$ individuals decreases to 1 as the carrying capacity is approached
*mutation occurs


```{r}
simulate_one_gen_mig_logistic <- function(N_a, N_A, carrying_capacity, decay_rate, sel_coeff, mut_rate, mig_rate_a) {
  # adult migrants arrive and are added to 'a' population
  N_a_mig <- rpois(1, mig_rate_a) # stochastic immigration
  N_a <- N_a + N_a_mig
  
  # determine total population size
  N_total <- N_a+N_A
  
  # reproduction occurs according to decay rates and selection coefficients
  offsp_a <- rpois(1, N_a * (1-decay_rate)) # independent of carrying capacity
  offsp_A <- rpois(1, N_A + N_A*(sel_coeff-decay_rate)*(1-N_total/carrying_capacity))
  # randomly kill individuals if carrying capacity is exceeded
  offsp_total <- offsp_A + offsp_a
  if(offsp_total > carrying_capacity)
  {
    offsp_a <- rbinom(1,carrying_capacity,c(offsp_a,offsp_A)/offsp_total)
    offsp_A <- carrying_capacity-offsp_a
  }

  
  # draw new mutants according to Poisson distribution
  mut_a_to_A <- rpois(1, offsp_a * mut_rate)
  
  # determine new population sizes of wild type and mutant
  N_a_new <- offsp_a - mut_a_to_A
  N_A_new <- offsp_A + mut_a_to_A
  
  return(c(N_a_new, N_A_new))
}
# Test the function
print(simulate_one_gen_mig_logistic(N_a=50, N_A=0, carrying_capacity=50, decay_rate=0.1, sel_coeff=0.3, mut_rate=0.1, mig_rate_a=1))

```

### Questions

-   Why do we substract `mut_a_to_A` from `N_a_new` and not from `N_A_new`?
-   In the life cycle as defined above, do individuals reproduce first and then mutate, or vice versa?
-   Running the test output repeatedly, why do you almost always get mutant individuals in the next generation, even though you start without any mutants?
-   What does the function `c` do?

## Simulate a population trajectory

We can now track the dynamics of the population from one generation to another, but this is not enough. We want to follow the population from time 0 until it either is rescued or goes extinct. This is what the following function does.

```{r}
simulate_pop_mig_logistic <- function(N_init_a, N_init_A, carrying_capacity, decay_rate, sel_coeff, mut_rate, mig_rate_a, t_max) {
  # Create the vector in which to save the results
  pop_vector <- c(N_init_a, N_init_A)
  # initiate the variables
  pop_new <- c(N_init_a, N_init_A)
  
  # run the simulation until generation t_max
  for (i in 1:t_max+1) {
    # redefine the current population one generation later
    pop_new <- simulate_one_gen_mig_logistic(pop_new[1],pop_new[2],carrying_capacity, decay_rate, sel_coeff, mut_rate, mig_rate_a)
    # add the new population sizes to the output vector
    pop_vector <- rbind(pop_vector,pop_new)
  }
  
  # define the row and column names of the output vector
  rownames(pop_vector) <- (0:t_max)[1:length(pop_vector[,1])] # note that the vector has to be cut if the simulation stopped early
  colnames(pop_vector) <- c("a","A")
  # return the result
  return(pop_vector)	
}


# Test the function and plot the result
# set t_max to a value to be used througout
max_gen <- 5000
# create your simulation data
output <- simulate_pop_mig_logistic(1000,0,1000,0.1,0.3,0.001,3,max_gen)
# show the last few lines of the data table
print(tail(output))
# plot the output - note that the plot range and the x axis have to be adjusted based on the length of the table
# determine x axis range 
x_range <- 0:(length(output[,1])-1)
# this plots the total population size
plot(x_range,output[,1]+output[,2],type='l',ylim=c(0,max(output[,1]+output[,2])),xlab = "Generation",ylab = "Population size")
# add number of wild type individuals
lines(x_range,output[,1], col="blue")
# add number of mutant individuals
lines(x_range,output[,2], col="red")
```

## What could be interesting analysis outputs from this model?

Important to note: extinction is impossible because new migrants always come in. But it seems that for high immigration rates and relatively low carrying capacity, the mutant can repeatedly get lost even if it is advantageous in the environment. This complicates the analysis.

### What could we measure?

- Time until the carrying capacity is reached. *This indicates that the mutant has taken over and "rescued" the population -- at least temporarily.*
- Average frequency of the two types over the last XXX (number to be determined, e.g. 100 or 1000?) generations of the simulation. *This describes how much the population is suffering from the inflow of deleterious genotypes.*
- Count whether or how often or how fast the mutant is lost again after "rescuing" the population. *This describes how sensitive the rescued population is to swamping from the upstream population.*

### Which parameters to vary, and which to keep constant?





## Find time to carrying capacity

We can now track the dynamics of the population from one generation to another, but this is not enough. We want to follow the population from time 0 until it either is rescued or goes extinct. This is what the following function does.

```{r}
simulate_pop_mig_logistic <- function(N_init_a, N_init_A, carrying_capacity, decay_rate, sel_coeff, mut_rate, mig_rate_a, t_max) {
  
  # initiate the variables
  pop_new <- c(N_init_a, N_init_A)
  # start counting generations
  gen <- 0
  
  # run the simulation until generation t_max
  for (i in 1:t_max+1) {
    # redefine the current population one generation later
    pop_new <- simulate_one_gen_mig_logistic(pop_new[1],pop_new[2],carrying_capacity, decay_rate, sel_coeff, mut_rate, mig_rate_a)
    # adding one generation to counter
    gen <- gen+1
    # calculate population size
    N_total <- sum(pop_new)
    # stop when carrying capacity is reached
    if (N_total == carrying_capacity) break
  }
  return(gen)	
}


# Test the function and plot the result
# set t_max to a value to be used througout
max_gen <- 100
# create your simulation data
output <- simulate_pop_mig_logistic(1000,0,1000,0.1,0.3,0.001,3,max_gen)
print(output)
```
